A Plasma Torus Around a Young Low-Mass Star

### 2025-03-17 paper/ms.tex
update

% photometry
Figure~\ref{fig:spec} shows the data from February 2024.  Over
timescales of years, CPVs maintain the same period, while their
photometric morphology evolves.  TIC~141146667 indeed evolved relative
to the February 2022 discovery data \cite{Bouma2024}.  In the February
2024 data, the average photometric signal showed a small brightening
over 45\% of the period, followed by a complex flux dip spanning 55\%
of the period.  This eclipse feature shows two to three local
photometric minima, and one to two local maxima.  Its W-shape is
suggestive of eclipse geometries seen in some forward models of warped
plasma tori \cite{Townsend2008}.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/sf1.pdf}
  \caption{Photometric evolution of TIC 141146667 near the epoch of
  spectroscopic observation (green bar). 
  {\bf Panel a}: TESS simple aperture photometry. 
  The main data gaps were caused by scattered light from the Earth
  (BTJD 3356-3358.5) and Moon (BTJD 3359.5-3360.5).
  {\bf Panels b,c}: Folded light curve before and after spectroscopy.
  Although some evolution in the light curve morphology occurred between
  the two epochs, the large, complex eclipse feature remained present.
  }
  \label{fig:fulllc}
\end{figure}

{\it Photometry:} TESS observed
TIC~141146667 ($T$=13.3) in Sectors 14, 15, 21, 41, 48, and 75.
Two-minute data were acquired during Sectors 41, 48 (TESS DDT039,
PI: Kunimoto), and 75 (TESS Program G06030, PI: Bouma).  The data
from Sectors 14, 15, and 21 had a 30-minute cadence, which
smears sharp features over the $<$4\,hour period
(see \cite{Gunther2022}).  The nearest known star, TIC~141146666
($T$=14.5), is 25$''$ from TIC~141146667 and is photometrically stable
in the pixel-level TESS data.

Figure~\ref{fig:fulllc} shows the data from Sector~75, which were
acquired using the camera that pointed closest to the ecliptic.  The
largest gap in coverage is from BTJD 3356.0~-~3358.5, and includes
the Keck/HIRES observation epoch (green bar).   There are no flux
measurements reported during this interval because the Earth was
within 25$^\circ$ the camera's boresight, yielding high levels of
scattered light.  From BTJD 3359.4~-~3362.0, the Moon then passed
within 25$^\circ$ of the camera's boresight.  Based on the observed
level of scattered light in the optimal TIC~141146667 aperture, we
manually masked out times from 3359.40~-~3360.13, and judged the
remainder of the data during the lunar approach to be usable.  The
small data gaps from BTJD 3353.55~-~3353.77 and from BTJD
3360.12~-~3360.33 were caused by data downlinks at the spacecraft's
perigee and apogee, respectively.

{\it Spectroscopy:}
We observed TIC~141146667 ($V$=16.2) with Keck/HIRES for five hours over a
second-half night spanning UT 2024-02-17 10:47 to UT 2024-02-17
16:13.  The star's airmass over this window spanned $z$=1.2-2.2, and
we opted for a fixed 15 minute cadence over the entire sequence,
except for a final 10 minute exposure due to increasing sky
brightness at sunrise.  We observed without the iodine cell and used
the C2 decker (0$\farcs$86$\times$14$\farcs$0) in the red instrument
configuration, yielding a spectral resolution $R$$\approx$45{,}0000
($\delta v$$\approx$6.7\,\kms; $\delta v / v_{\rm
eq}$$\approx$0.05).  We binned the CCD readout by a factor of three
in the spatial dimension, yielding $\approx$1,000 photons (S/N=33)
per pixel in the continuum at 6500\,\AA, at minimum airmass.  Strong
winds contributed to 1\farcs2$\pm$0\farcs2 seeing over the night,
but conditions were otherwise favorable.  We reduced the
echelleogram to a one-dimensional spectrum using the standard
techniques of the California Planet Survey \cite{Howard2010}.
Figure~\ref{fig:spec}b shows the result in the vicinity of H$\alpha$
without any additional processing.

@@

{\it Age: Lithium}---The depletion of lithium due to $^7$Li(p,
$\alpha$)$^4$He burning in the cores of low-mass stars has been studied
for over sixty years \cite{Hayashi1963,Bildsten1997,Burke2004}.
\cite{Wood2023} provided an instructive recent overview: an abbreviated
summary is that cooler and younger M dwarfs can show the 6707.8\,\AA\
doublet in absorption, $\gtrsim$10\% below their ``continua''.  Unlike
Sun-like stars, the continuum for M dwarfs is poorly defined due to
their molecular absorption.  We attempted a lithium measurement by
constructing a wavelength-binned and Doppler-corrected TIC~141146667
spectrum, assigning its uncertainties based on the measured scatter
across the five hour dataset.  We then compared this average spectrum
against the nearest matching template from \cite{Bochanski2007}.  The
data show a small depression near the expected lithium wavelength,
potentially consistent with a $\Delta \lambda$$\approx$3\,\AA\ line
broadening.  This feature nominally yields ${\rm EW}_{\rm
Li}$=71$^{+18}_{13}$\,m\AA, where the statistical uncertainties are
evaluated using a bootstrap resampling technique from the statistical
uncertainties in the spectrum.  The systematic uncertainties associated
with the continuum normalization are larger, and likely comparable to
the amplitude of the feature; we therefore treat the result of this
measurement as a $2\sigma$ upper limit: ${\rm EW}_{\rm Li}$$<$107\,m\AA.

Despite uncertainties in the details, what can be stated with confidence
is that lithium is not abundant in the spectrum of TIC~141146667.
Figure~\ref{fig:liew_population} compares our upper limit against the
equivalent width measurements reported by \cite{Jeffries2023}
based on the Gaia-ESO spectroscopic survey.  If the star were
$\lesssim$20\,Myr old, at its temperature we would expect to see
lithium in abundance ($>$400\,m\AA).  Since we do not, we can set an
empirical bound on the lithium-derived age of $t_{\rm
Li,emp}$$\gtrsim$20\,Myr.  The \cite{Feiden2016} lithium isochrones
provide a point for theoretical comparison, and suggest that since
$M_{\rm K}$$=$6.67\,mag, $t_{\rm Li,th}$$\gtrsim$35\,Myr is the
theoretical age at which complete depletion occurs in a star with this
luminosity (see e.g.~Figure 7 from \cite{Wood2023}).

@@

How much dust is allowed, given the lack of an infrared excess?
Figure~\ref{fig:sed} shows the SED.  While AllWISE \cite{Cutri2014}
yielded a confident W3 detection (9.8$\sigma$), the W4 extraction
yielded only a marginal (1.7$\sigma$) indication of detectable flux.
Similar to other CPVs \cite{Stauffer2017,Bouma2024}, the photometric
uncertainties from WISE W1 and W2 allow at most a $\lesssim$2\% excess
at 3-5\,$\mu$m relative to the stellar photosphere, and a $\lesssim$5\%
excess at 10\,$\mu$m (W3).  To estimate the implied mass bound, we
assume $T_{\rm dust}$=1500\,K, typical for dust near the star (see
\cite{Zhan2019} for discussion regarding dust sublimation).  We then
treat emission from the dust and the star as Planck functions,
requiring $F_{\lambda,\star} > f \cdot F_{\lambda,\mathrm{dust}}$, for
$f$ the scaling factor set by the WISE photometric uncertainty.  For
this condition to hold, we find that the bolometric flux ratio $F_{\rm
dust}/F_{\star}$$<$6$\cdot$10$^{-3}$.  Scaling the associated
luminosities ($L = \sigma_{\rm SB} T_{\rm eff}^4 / (4\pi R^2)$)
implies an emitting area for the dust $<$10\% of the star's surface
area.  Converting this constraint to a dust mass requires an
assumption regarding the grain properties; assuming a grain density
$\rho_{\rm d}$=3\,g\,cm$^{-3}$ typical for silicate grains, and a fixed grain
size $a$=1\,$\mu$m \cite{Tanimoto2020,Gunther2022,Koen2023} yields an
upper limit of $M_{\rm dust}\lesssim$10$^{17}$\,g.

@@

>>>>

% photometry
Figure~\ref{fig:spec} shows the data from February 2024.  Over
timescales of years, CPVs maintain a fixed period while their
photometric morphology evolves.  TIC~141146667 indeed evolved relative
to the February 2022 discovery data \cite{Bouma2024}.  In the February
2024 TESS data, the average photometric signal showed a small brightening
over 45\% of the period, followed by a complex flux dip spanning 55\%
of the period.  This eclipse feature shows two to three local
photometric minima, and one to two local maxima.  Its W-shape is
suggestive of eclipse geometries seen in some forward models of warped
plasma tori \cite{Townsend2008}.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/sf1.pdf}
  \caption{Photometric evolution of TIC 141146667 near the Keck/HIRES observation (green bar). 
  {\bf Panel a}: TESS simple aperture photometry. 
  The main data gaps were caused by scattered light from the Earth
  (BTJD 3356-3358.5) and Moon (BTJD 3359.5-3360.5).
  Raw two minute data are in gray; black time-averages ten minute
  sampling.
  {\bf Panels b,c}: Folded light curve before and after spectroscopy.
  Raw two minute data are in gray; black phase-averages to 100 points
  per cycle.
  In the BTJD 3352-3356 window, a state switch occurred near BTJD 3353,
  in which the dip at $\phi$$\approx$0.8 disappeared.
  While the large, complex eclipse feature remained present before and
  after the HIRES sequence, its photometric shape evolved during the
  data gap.
  }
  \label{fig:fulllc}
\end{figure}

{\it Photometry:} TESS observed TIC~141146667 ($T$=13.3) in Sectors 14,
15, 21, 41, 48, and 75.  Two-minute data were acquired during Sectors
41, 48 (TESS DDT039, PI: Kunimoto), and 75 (TESS Program G06030, PI:
Bouma).  The data from Sectors 14, 15, and 21 had a 30-minute cadence,
which smears sharp features over the $<$4\,hour period (see
\cite{Gunther2022}).  The field is not crowded:  the nearest known
star, TIC~141146666 ($T$=14.5), is 25$''$ from TIC~141146667 and is
photometrically stable in the pixel-level TESS data.

Figure~\ref{fig:fulllc} shows the TESS Sector~75 data acquired near
the epoch of spectroscopy.  The gap in coverage from BTJD
3356.0~-~3358.5 occurred because the Earth was within 25$^\circ$ of
the TESS camera's boresight, yielding high levels of scattered light;
this gap included the Keck/HIRES observation epoch (green bar).  From
BTJD 3359.4~-~3362.0, the Moon then passed within 25$^\circ$ of the
camera's boresight.  Based on the observed level of scattered light in
the optimal TIC~141146667 aperture, we manually masked out times from
3359.40~-~3360.13, and judged the remainder of the data during the
lunar approach to be usable.  Small data gaps from BTJD
3353.55~-~3353.77 and from BTJD 3360.12~-~3360.33 were caused by data
downlinks at the spacecraft's perigee and apogee, respectively.

Broadly, Figure~\ref{fig:fulllc} shows that a large, complex eclipse
feature was present before and after the HIRES data were
acquired.  From BTJD 3352~-~3356, this eclipse had two sharp local
minima;  the local minimum at $\phi$$\approx$0.8 disappeared following
the flare at BTJD 3353, yielding an eclipse more closely resembling a
long ``V'' than a ``W''.  Similar state changes have previously been
noted and discussed \cite{Stauffer2017,Bouma2024}.  The photometric
shape therefore evolved during the twelve cycles spanning the
3356~-~3358 gap, since the average shape from 3358-3363 more
closely resembles the initial ``W''.

{\it Spectroscopy:}
We observed TIC~141146667 ($V$=16.2) with Keck/HIRES for five hours over
a second-half night spanning 17 February 2024 10:47 to 16:13 (UT).
The star's airmass over this window spanned $z$=1.2-2.2, and we opted
for a fixed 15 minute cadence over the entire sequence, except for a
final 10 minute exposure due to increasing sky brightness at sunrise.
We observed without the iodine cell and used the C2 decker
(0$\farcs$86$\times$14$\farcs$0) in the red instrument configuration,
yielding a spectral resolution $R$$\approx$45{,}0000 ($\delta
v$$\approx$6.7\,\kms; $\delta v / v_{\rm eq}$$\approx$0.05).  We binned
the CCD readout by a factor of three in the spatial dimension, yielding
$\approx$1,000 photons (S/N=33) per pixel in the continuum at 6500\,\AA,
at minimum airmass.  Strong winds contributed to 1\farcs2$\pm$0\farcs2
seeing over the night, but conditions were otherwise favorable.  We
reduced the echelleogram to a one-dimensional spectrum using the
standard techniques of the California Planet Survey \cite{Howard2010}.
Figure~\ref{fig:spec}b shows the result in the vicinity of H$\alpha$
without any additional processing.

@@

{\it Age: Lithium}---The depletion of lithium due to $^7$Li(p,
$\alpha$)$^4$He burning in the cores of low-mass stars has been studied
for over sixty years \cite{Hayashi1963,Bildsten1997,Burke2004}.
\cite{Wood2023} provided an instructive recent overview: an abbreviated
summary is that sufficiently cool and young M dwarfs show the 6707.8\,\AA\
doublet in absorption, $\gtrsim$10\% below their ``continua''.  Unlike
for Sun-like stars, the continuum for M dwarfs is poorly defined due to
their molecular absorption.  We attempted a lithium measurement by
constructing a wavelength-binned and Doppler-corrected TIC~141146667
spectrum, assigning its uncertainties based on the measured scatter
across the five hour dataset.  We then compared this average spectrum
against the nearest matching template from \cite{Bochanski2007}.  The
data show a small depression near the expected lithium wavelength,
potentially consistent with the $\Delta \lambda$$\approx$3\,\AA\ line
broadening.  This feature nominally yields ${\rm EW}_{\rm
Li}$=71$^{+18}_{-13}$\,m\AA, where the statistical uncertainties are
evaluated using a bootstrap resampling technique from the statistical
uncertainties in the spectrum.  However the systematic uncertainties associated
with the continuum normalization are larger, and likely comparable to
the amplitude of the feature; we therefore treat the result of this
measurement as a $2\sigma$ upper limit: ${\rm EW}_{\rm Li}$$<$107\,m\AA.

Despite uncertainties in the details, what can be stated with confidence
is that lithium is not abundant in the spectrum of TIC~141146667.
Figure~\ref{fig:liew_population} compares our upper limit against the
equivalent width measurements reported by \cite{Jeffries2023}
based on the Gaia-ESO spectroscopic survey.  If the star were
$\lesssim$20\,Myr old, at its temperature we would expect to see
lithium in abundance ($>$400\,m\AA).  Since we do not, we can set an
empirical bound on the lithium-derived age of $t_{\rm
Li,emp}$$\gtrsim$20\,Myr.  The \cite{Feiden2016} lithium isochrones
provide a point for theoretical comparison, and suggest that since
$M_{\rm K}$$\approx$6.67\,mag, $t_{\rm Li,th}$$\gtrsim$35\,Myr is the
theoretical age at which complete depletion occurs in a star with this
luminosity (see e.g.~Figure 7 from \cite{Wood2023}).

@@

How much dust is allowed, given the lack of an infrared excess?
Figure~\ref{fig:sed} shows the SED.  While AllWISE \cite{Cutri2014}
yielded a confident W3 detection (9.8$\sigma$), the W4 extraction
yielded only a marginal indication (1.7$\sigma$) of detectable flux.
Similar to other CPVs \cite{Stauffer2017,Bouma2024}, the photometric
uncertainties from WISE W1 and W2 allow at most a $\lesssim$2\% excess
at 3-5\,$\mu$m relative to the stellar photosphere, and a $\lesssim$5\%
excess at 10\,$\mu$m (W3).  To estimate the implied mass bound, we
assume $T_{\rm dust}$=1500\,K, typical for dust near the star (see
\cite{Zhan2019} for discussion regarding dust sublimation).  We then
treat emission from the dust and the star as Planck functions,
requiring $F_{\lambda,\star} > f \cdot F_{\lambda,\mathrm{dust}}$, for
$f$ the scaling factor set by the WISE photometric uncertainty.  For
this condition to hold, we find that the bolometric flux ratio $F_{\rm
dust}/F_{\star}$$<$6$\cdot$10$^{-3}$.  Scaling the associated
luminosities ($L = \sigma_{\rm SB} T_{\rm eff}^4 / (4\pi R^2)$)
implies an emitting area for the dust $<$10\% of the star's surface
area.  Converting this constraint to a dust mass requires an
assumption regarding the grain properties; assuming a grain density
$\rho_{\rm d}$=3\,g\,cm$^{-3}$ typical for silicate grains, and a fixed grain
size $a$=1\,$\mu$m \cite{Tanimoto2020,Gunther2022,Koen2023} yields an
upper limit of $M_{\rm dust}\lesssim$10$^{17}$\,g.

@@

Ha
Hgamma
Hdelta?
He5875
CaK
K7700
Li6708
